---
title: "Homework 04"
author: "Spencer Pease"
date: "2/5/2020"
output: pdf_document
---

```{r setup, include=FALSE}

# Set output options
knitr::opts_chunk$set(echo = FALSE)
options(knitr.kable.NA = '-')

# Include libraries
library(dplyr)
library(readr)
library(ggplot2)
library(uwIntroStats)

# Helper functions
source("functions/summary_table.R")
source("functions/inference_table.R")

# Prepare data
smsa <- read_csv("data/SMSA.csv") %>%
  select(mort = Mortality, nox = NOx, rain = Rain) %>%
  mutate(
    log_nox = log(nox),
    low_rain = factor(rain < 20, levels = c(TRUE, FALSE), labels = c("Low Rain*", "Other"))
  )

mri2 <- read_csv("data/MRI2.csv") %>%
  select(height, weight, male) %>%
  mutate(sex = factor(male, levels = c(0, 1), labels = c("Female", "Male")))

```

# _(Q1)_

```{r}
mort_nox_scatter <- ggplot(smsa, aes(x = log_nox, y = mort)) +
  labs(
    title = "Mortality vs NOx Level",
    x = "log(NOx)",
    y = "Mortality"
  ) +
  theme_bw() +
  theme(
    text = element_text(family = "serif"),
    legend.position = "top"
  )

```

```{r}
mort_nox_scatter + geom_point()
```

The scatter of _mortality_ vs $log(NOx)$ shows at least a small positive
first-order trend. There are four locations however that shun this trend and
maintain lower mortality for comparatively high $log(NOx)$ levels.

# _(Q2)_

Regression model:

$$ E[mortality | log(NOx)] = \beta_{0} + \beta_{1} \times log(NOx) $$

```{r}
mort_nox_lm <- regress("mean", mort ~ log_nox, data = smsa)

mort_nox_inf_tbl <- inference_table(mort_nox_lm, paramter_names = "log(NOx)")

knitr::kable(mort_nox_inf_tbl, booktabs = TRUE, digits = 3,
  caption = "Inferential statistics of Mortality vs log(NOx)"
)

q2 <- list(
  round(mort_nox_inf_tbl[[1, 2]], 2),
  round(mort_nox_inf_tbl[[1, 4]], 2),
  round(mort_nox_inf_tbl[[1, 5]], 2),
  round(mort_nox_inf_tbl[[1, 7]], 2)
)

```

Using the above linear regression model with robust standard error estimates to
assess the first-order relationship between mortality and $log(NOx)$, we
estimate that an increase in one unit $log(NOx)$ is associated with a `r q2[[1]]`
unit increase in mortality ($95\%$ CI: `r q2[[2]]`, `r q2[[3]]`). A _P_-value
greater than $.05$ suggests this relationship is not statistically significant.

# _(Q3)_

Using robust standard error estimates avoids the assumption of equal variance
within each group of our data (homoscedasticity), meaning the model makes fewer
assumptions about the data in general.

This model is still based on the assumptions that the data are independent and
the sample size is sufficiently large for the central limit theorem to apply when
testing associations. The point estimate and confidence interval also rely on
the assumption that the model is a good fit to the true distribution.

# _(Q4)_

```{r}
mort_nox_scatter +
  geom_point(aes(color = low_rain)) +
  labs(caption = "*Low Rain: < 20 inches of rain per year") +
  scale_color_manual(name = NULL, values = c("salmon", "black"))
```

Rainfall should be a confounder, since it appears to have an association with
both _NOx_ pollution level and mortality. It doesn't make sense for pollution
to determine rainfall, so rain is not in the casual pathway of interest.

# _(Q5)_

Regression model:

$$ E[mortality | log(NOx), rain] = \beta_{0} + \beta_{1} \times log(NOx) + \beta_{2} \times rain + \beta_{3} \times log(NOx) \cdot rain $$

```{r}
mort_nox_rain_lm <- regress("mean", mort ~ log_nox + rain + log_nox * rain, data = smsa)

mort_nox_rain_inf_tbl <- mort_nox_rain_lm %>%
  inference_table(paramter_names = c("log(NOx)", "Rain", "Log(NOx):Rain")) %>%
  mutate(`Pr(>|t|)` = format(`Pr(>|t|)`, digits = 3))

knitr::kable(mort_nox_rain_inf_tbl, booktabs = TRUE, digits = 3,
  caption = "Inferential statistics of Mortality vs log(NOx), adjusted for Rain"
)

```

# _(Q6)_



# _(Q7)_

# _(Q8)_

## _(Q8.a)_

```{r}
weight_height_scatter <- ggplot(mri2, aes(x = height, y = weight)) +
  geom_point(aes(color = sex), shape = 1, alpha = .4) +
  stat_smooth(aes(color = sex), method = "lm", se = FALSE) +
  stat_smooth(method = "lm", se = FALSE, linetype = "dashed", color = "black") +
  labs(
    title = "Weight vs Height",
    x = "Height (cm)",
    y = "Weight (lbs)",
    caption = "Colored sex-specific regression lines; dashed both-sex regression line",
    color = NULL
  ) +
  theme_bw() +
  theme(
    text = element_text(family = "serif"),
    legend.position = "top"
  )
```

```{r}
weight_height_scatter
```

```{r}

weight_height_lms <- list(
  both = regress("mean", weight ~ height, data = mri2),
  male = regress("mean", weight ~ height, data = filter(mri2, male == 1)),
  female = regress("mean", weight ~ height, data = filter(mri2, male == 0))
)

weight_height_comb_inf_tbl <- weight_height_lms %>%
  purrr::imap_dfr(~inference_table(lm = .x, paramter_names = .y)) %>%
  mutate(
    `Pr(>|t|)` = format(`Pr(>|t|)`, digits = 3),
    Data = Parameter,
    Parameter = "Height"
  ) %>%
  select(Data, everything())

knitr::kable(weight_height_comb_inf_tbl, booktabs = TRUE, digits = 3,
  caption = paste("Inferential statistics of Weight vs Height for separate models",
                  "containing both sexes and individual sex data subsets")
)

```

## _(Q8.b)_



## _(Q8.c)_
## _(Q8.d)_
## _(Q8.e)_
## _(Q8.f)_

```{r}
weight_height_sex_lm <- regress("mean", weight ~ height + male, data = mri2)

weight_height_sex_inf_tbl <- weight_height_sex_lm %>%
  inference_table() %>%
  mutate(`Pr(>|t|)` = format(`Pr(>|t|)`, digits = 3))

knitr::kable(weight_height_sex_inf_tbl, booktabs = TRUE, digits = 3,
  caption = "Inferential statistics of Weight vs Height, adjusted for Sex"
)

```


# _(Q9)_
