---
title: "Homework 04"
author: "Spencer Pease"
date: "2/5/2020"
output: pdf_document
---

```{r setup, include=FALSE}

# Set output options
knitr::opts_chunk$set(echo = FALSE)
options(knitr.kable.NA = '-')

# Include libraries
library(dplyr)
library(readr)
library(ggplot2)
library(uwIntroStats)

# Helper functions
source("functions/summary_table.R")
source("functions/inference_table.R")

# Prepare data
smsa <- read_csv("data/SMSA.csv") %>%
  select(mort = Mortality, nox = NOx, rain = Rain) %>%
  mutate(
    log_nox = log(nox),
    low_rain = factor(rain < 20, levels = c(TRUE, FALSE), labels = c("Low Rain*", "Other"))
  )

mri2 <- read_csv("data/MRI2.csv") %>%
  select(height, weight, male) %>%
  mutate(sex = factor(male, levels = c(0, 1), labels = c("Female", "Male")))

```

# _(Q1)_

```{r}
mort_nox_scatter <- ggplot(smsa, aes(x = log_nox, y = mort)) +
  labs(
    title = "Mortality vs NOx Level",
    x = "log(NOx)",
    y = "Mortality"
  ) +
  theme_bw() +
  theme(
    text = element_text(family = "serif"),
    legend.position = "top"
  )

```

```{r}
mort_nox_scatter + geom_point()
```

The scatter of _mortality_ vs $log(NOx)$ shows at least a small positive
first-order trend. There are four locations however that shun this trend and
maintain lower mortality for comparatively high $log(NOx)$ levels.

# _(Q2)_

Regression model:

$$ E[mortality | log(NOx)] = \beta_{0} + \beta_{1} \times log(NOx) $$

```{r}
mort_nox_lm <- regress("mean", mort ~ log_nox, data = smsa)

mort_nox_inf_tbl <- inference_table(mort_nox_lm, paramter_names = "log(NOx)")

knitr::kable(mort_nox_inf_tbl, booktabs = TRUE, digits = 3,
  caption = "Inferential statistics of Mortality vs log(NOx)"
)

q2 <- list(
  round(mort_nox_inf_tbl[[1, 2]], 2),
  round(mort_nox_inf_tbl[[1, 4]], 2),
  round(mort_nox_inf_tbl[[1, 5]], 2),
  round(mort_nox_inf_tbl[[1, 7]], 2)
)

```

Using the above linear regression model with robust standard error estimates to
assess the first-order relationship between mortality and $log(NOx)$, we
estimate that an increase in one unit $log(NOx)$ is associated with a `r q2[[1]]`
unit increase in mortality ($95\%$ CI: `r q2[[2]]`, `r q2[[3]]`). A _P_-value
greater than $.05$ suggests this relationship is not statistically significant.

# _(Q3)_

Using robust standard error estimates avoids the assumption of equal variance
within each group of our data (homoscedasticity), meaning the model makes fewer
assumptions about the data in general.

This model is still based on the assumptions that the data are independent and
the sample size is sufficiently large for the central limit theorem to apply when
testing associations. The point estimate and confidence interval also rely on
the assumption that the model is a good fit to the true distribution.

# _(Q4)_

```{r}
mort_nox_scatter +
  geom_point(aes(color = low_rain)) +
  labs(caption = "*Low Rain: < 20 inches of rain per year") +
  scale_color_manual(name = NULL, values = c("salmon", "black"))
```

Rainfall should be an effect modifier, since it appears to have an association
with both _NOx_ pollution level and mortality. It doesn't make sense for
pollution to determine rainfall, so rain is not in the casual pathway of
interest.

# _(Q5)_

Regression model:

$$ E[mortality | log(NOx), rain] = \beta_{0} + \beta_{1} \times log(NOx) + \beta_{2} \times rain + \beta_{3} \times log(NOx) \cdot rain $$

```{r}
mort_nox_rain_lm <- regress("mean", mort ~ log_nox + rain + log_nox * rain, data = smsa)

mort_nox_rain_inf_tbl <- mort_nox_rain_lm %>%
  inference_table(paramter_names = c("log(NOx)", "Rain", "Log(NOx):Rain")) %>%
  mutate(`Pr(>|t|)` = format(`Pr(>|t|)`, digits = 3))

knitr::kable(mort_nox_rain_inf_tbl, booktabs = TRUE, digits = 3,
  caption = "Inferential statistics of Mortality vs log(NOx), adjusted for Rain"
)

q6 <- mort_nox_rain_inf_tbl[, c(2, 4, 5, 7)]
q6[, 1:3] <- round(q6[, 1:3], 2)

```

# _(Q6)_

Fitting the model displayed in _Q5_ using robust standard error estimates, we
produce the estimated coefficients in the above table. From these values we
estimate that for a location with no rainfall, one unit difference in $log(NOx)$
will change mortality by `r q6[1, 1]` units ($95\%$ CI: `r q6[1, 2:3]`).
Between two groups with a one inch per year difference in rainfall and equal
$log(NOx)$, mortality will in change by `r q6[2, 1]` units
($95\%$ CI: `r q6[2, 2:3]`). However, when comparing two groups that differ by
one unit in both rainfall per year and $log(NOx)$, the resulting change in
difference in mortality is `r sum(q6[, 1])` units, Which is less negative than
just the sum of the $log(NOx)$ and rain coefficients. Of all of these
estimates, rainfall on its own is not statistically significant. Overall, the
relationship between mortality and $log(NOx)$ tends to be negative unless there
is a significant difference in amount of rainfall (approximately 16 inches per
year) between two groups.

# _(Q7)_

```{r}

aridia_coef <- NULL

```

In Aridia, less rain means that the negative impact of $NOx$ level on mortality
is diminished. This means reducing pollution might not be as effective at
lowering mortality as other locations with more rain.

In Seattle, large amounts of rain increase the impact of $NOx$ on mortality, so
lowering pollution will have a larger impact on mortality.


# _(Q8)_

## _(Q8.a)_

```{r}
weight_height_scatter <- ggplot(mri2, aes(x = height, y = weight)) +
  geom_point(aes(color = sex), shape = 1, alpha = .4) +
  stat_smooth(aes(color = sex), method = "lm", se = FALSE) +
  stat_smooth(method = "lm", se = FALSE, linetype = "dashed", color = "black") +
  labs(
    title = "Weight vs Height",
    x = "Height (cm)",
    y = "Weight (lbs)",
    caption = "Colored sex-specific regression lines; dashed both-sex regression line",
    color = NULL
  ) +
  theme_bw() +
  theme(
    text = element_text(family = "serif"),
    legend.position = "top"
  )
```

```{r}
weight_height_scatter
```

```{r}

weight_height_lms <- list(
  both = regress("mean", weight ~ height, data = mri2),
  male = regress("mean", weight ~ height, data = filter(mri2, male == 1)),
  female = regress("mean", weight ~ height, data = filter(mri2, male == 0))
)

weight_height_comb_inf_tbl <- weight_height_lms %>%
  purrr::imap_dfr(~inference_table(lm = .x, paramter_names = .y)) %>%
  mutate(
    `Pr(>|t|)` = format(`Pr(>|t|)`, digits = 3),
    Data = Parameter,
    Parameter = "Height"
  ) %>%
  select(Data, everything())

knitr::kable(weight_height_comb_inf_tbl, booktabs = TRUE, digits = 3,
  caption = paste("Inferential statistics of Weight vs Height for separate models",
                  "containing both sexes and individual sex data subsets")
)

```

## _(Q8.b)_

Scientifically, sex is an important variable to consider when predicting weight
from height. It is well documented that a person's sex plays a role in their
expected height and weight. Regardless if sex has a statistically significant
role in our sample, our model should reflect what we expect in the true
population. To exclude sex from our model would be to ignore a documented
phenomena in our population.

## _(Q8.c)_

The estimated slopes from the sex-specific models are nearly identical,
suggesting that the relationship between weight and height is similar for both
males and females.

```{r}
height_sex_tbl <- mri2 %>%
  select(male, height) %>%
  group_by(male) %>%
  summary_table()

knitr::kable(height_sex_tbl, booktabs = TRUE, digits = 3,
  caption = "Summary of distribution of heights by sex"
)
```

From the data in our sample, we see that males have a greater height than
females on average. This means there is likely an interaction between height
and sex.

## _(Q8.d)_

Altogether, sex plays the role of a confounder.

## _(Q8.e)_

The estimated slope of the both-sex regression is slightly greater than both of
the estimated sex-specific regression slopes. Since the both-sex regression
includes both the larger heights and weights of the males and the lower heights
and weights of the females, the model $\beta_{1}$ will represent this more
extreme range of data as a greater estimated slope.

## _(Q8.f)_

```{r}
weight_height_sex_lm <- regress("mean", weight ~ height + male, data = mri2)

weight_height_sex_inf_tbl <- weight_height_sex_lm %>%
  inference_table() %>%
  mutate(`Pr(>|t|)` = format(`Pr(>|t|)`, digits = 3))

knitr::kable(weight_height_sex_inf_tbl, booktabs = TRUE, digits = 3,
  caption = "Inferential statistics of Weight vs Height, adjusted for Sex"
)

```

The estimated slope of the multiple regression model nearly matches the
estimated slopes of the sex specific models. This is what we would expect, since
adjusting for sex allows the model to capture separately the range of height
and weight values associated with each sex while maintaining the same
relationship between the two variables across sexes.

# _(Q9)_

The denominator of (4.4) includes the term $(1 - r^{2}_{X,W})$, which will
increase in magnitude (and thereby decrease $var(\hat{\gamma_{1}})$) as the
correlation between $X$ and $W$ decreases. In this scenario $X$ is the treatment
and $W$ is diabetes, which are uncorrelated with each other. This minimizes
$r^{2}_{X,W}$, which decreases the variance of our estimated coefficient
overall, creating a more efficient model for estimation.

Checking for an interaction between the treatment and diabetes in this analysis
is a good idea, because if there is an interaction between the two then it
would change the scientific question we should be asking since we would know
more about the behavior of our population.
