---
title: "Homework 06"
author: "Spencer Pease"
date: "2/28/2020"
output: pdf_document
---

```{r setup, include=FALSE}

# Set output options
knitr::opts_chunk$set(echo = FALSE)
options(knitr.kable.NA = '-')

# Include libraries
library(dplyr)
library(readr)
library(ggplot2)
library(uwIntroStats)
library(aod)

# Helper functions
source("functions/summary_table.R")
source("functions/inference_table.R")

# Prepare data
wcgs <- read_csv("data/wcgs.csv") %>%
  transmute(
    age,
    chd = if_else(chd69 == "no", 0, 1),
    type = factor(dibpat)
  )

```

# _(Q1)_ Association between age and CHD events

```{r}
chd_age_lrm <- regress("odds", chd ~ age, data = wcgs)
chd_age_coef <- chd_age_lrm$fit$coefficients
```


## _(Q1.a)_

Fitted logistic regression model with _having a CHD event_ as the binary
response and _age_ and the predictor:

$$
logit(P(chd = 1 \mid age)) = \beta_0 + \beta_1 \cdot age
$$

```{r}
chd_age_inf <- chd_age_lrm %>%
  inference_table(include_intercept = TRUE) %>%
  mutate(`Pr(>|z|)` = format(`Pr(>|z|)`, digits = 3))

q1a <- list(
  chd_age_inf[1, 4:6],
  chd_age_inf[2, 4:6]
) %>% purrr::map(as.numeric) %>% purrr::map(round, digits = 2)

knitr::kable(
  chd_age_inf, booktabs = TRUE, digits = 3,
  caption = "Inferential statistics of log-odds CHD event vs age"
)

```

From the fitted model, we estimate that on average the odds of having a CHD
event within five years of the study start date for the subset of the population
zero years old ($e^{\beta_0}$) is `r q1a[[1]][1]`
(95% CI: `r q1a[[1]][2]`, `r q1a[[1]][3]`). This value is nonsensical, since it
is far outside the range of observed data and we have no reason to think the
relationship between middle-aged men and CHD events would apply to children.
Comparing two groups differing by one year in age, on average the odds ratio
$e^{\beta_1}$  of experiencing a CHD event within five years of the study start
date is `r q1a[[2]][1]` (95% CI: `r q1a[[2]][2]`, `r q1a[[2]][3]`). This is of
scientific interest for testing for an association between age and CHD events in
the population within the range of observed data.

## _(Q1.b,c)_

```{r}
wcgs_aug <- wcgs %>%
  mutate(
    log_odds_chd = chd_age_coef[1] + age * chd_age_coef[2],
    prob_chd = exp(log_odds_chd) / (1 + exp(log_odds_chd))
  )

chd_age_plot_lo <- ggplot(wcgs_aug, aes(x = age, y = log_odds_chd)) +
  geom_point() +
  labs(
    title = "Log-Odds CHD event vs Age",
    x = "Age (years)",
    y = "Log-Odds CHD event"
  ) +
  theme_bw() +
  theme(text = element_text(family = "serif"))

chd_age_plot_pr <- ggplot(wcgs_aug, aes(x = age, y = prob_chd)) +
  geom_point() +
  labs(
    title = "Probability of CHD event vs Age",
    x = "Age (years)",
    y = "Probability of CHD event"
  ) +
  theme_bw() +
  theme(text = element_text(family = "serif"))
```

```{r fig.height=4.2}
chd_age_plot_lo
chd_age_plot_pr
```

## _(Q1.d)_

```{r}
chd_age_or_rr_tbl <- tibble(
  Age = seq(40, 60, 5),
  log_odds = chd_age_coef[1] + Age * chd_age_coef[2],
  Odds = exp(log_odds),
  `Prob.` = Odds / (1 + Odds),
  OR = Odds / lag(Odds),
  RR = `Prob.` / lag(`Prob.`)
)
```

```{r}
chd_age_or_rr_tbl %>%
  select(Age, `Prob.`, Odds, RR, OR) %>%
  knitr::kable(
    booktabs = TRUE, digits = 2,
    caption = "Calculated risk ratio (RR) and odds ratio (OR)"
  )
```

From the table, we see that both the probability and odds of having a CHD event
are estimated to be higher on average for older ages. When looking at the
risk ratio (RR) and odds ratio (OR) between ages, we see that RR is lower
between older ages when the difference in age is constant, while the OR is the
same between any two ages of equal separation.

## _(Q1.e)_

```{r}
crit_val <- qnorm((1 - .95) / 2, lower.tail = FALSE)
chd_age_or5 <- (chd_age_inf[2, 2:3] * 5) %>%
  mutate(
    `95%L` = `Estimate` - crit_val * `Robust SE`,
    `95%H` = `Estimate` + crit_val * `Robust SE`
  ) %>%
  select(-`Robust SE`) %>%
  exp()
```


$$
OR_5 = \frac{Odds_{x+5}}{Odds_x}
= \frac{e^{\beta_0} e^{\beta_1 (x+5)}}{e^{\beta_0} e^{\beta_1 (x)}}
= e^{\beta_1 (x+5) - \beta_1 (x)}
= e^{\beta_1 (5)}
$$

```{r}
q1e <- list(chd_age_or5) %>%
  purrr::map(as.numeric) %>%
  purrr::map(round, digits = 2)

knitr::kable(
  chd_age_or5, booktabs = TRUE, digits = 2,
  caption = "Odds ratio comparing men 5 years apart in age"
)
```

The estimated average odds ratio between two subsets of the population five
years apart is `r q1e[[1]][1]` (95% CI: `r q1e[[1]][2]`, `r q1e[[1]][3]`). This
odds ratio matches the calculated odds ratios from (Q1.d), confirming that the
odds ratio is constant for a given interval between two groups, regardless of
the absolute value of the groups.

## _(Q1.f)_

```{r}
chd_age_inf[2, c(2:3, 7)]
```


Performing a Wald test on our logistic regression model fit with robust standard
error estimates


## _(Q1.g)_

```{r}
lrt_test <- aod::anova(glm(chd~1, data = wcgs), glm(chd~age, data = wcgs), test = "LRT")
```


# _(Q2)_ Association between personality and CHD events

## _(Q2.a)_

```{r}
chd_type_tbl <- xtabs(~ chd + type, data = wcgs) %>%
  addmargins() %>%
  as_tibble() %>%
  tidyr::pivot_wider(names_from = "type", values_from = "n")

chd_type_tbl %>%
  knitr::kable(
    format = "latex", booktabs = TRUE,
    caption = "Contingency table of CHD events and behavior types"
  ) %>%
  kableExtra::kable_styling(position = "center") %>%
  kableExtra::column_spec(c(1, 3), border_right = TRUE) %>%
  kableExtra::row_spec(2, hline_after = TRUE)
```


```{r}
chisq.test(wcgs$chd, wcgs$type) %>%
  broom::tidy() %>%
  select(`X-squared` = statistic, `P-value` = p.value, df = parameter) %>%
  mutate(`P-value` = format(`P-value`, digits = 3)) %>%
  knitr::kable(
    format = "latex", booktabs = TRUE, digits = 3,
    caption = "Chi-square test of CHD events and behavior types"
  )
```


 - $P(chd \mid type=A)$ : $\frac{178}{1589} =$ `r round(178/1589, digits = 2)`
 - $P(chd \mid type=B)$ : $\frac{79}{1565} =$ `r round(79/1565, digits = 2)`
 - $odds(chd \mid type=A)$ : $\frac{178}{1411} =$ `r round(178/1411, digits = 2)`
 - $odds(chd \mid type=B)$ : $\frac{79}{1486} =$ `r round(79/1486, digits = 2)`
 - $OR$ : $\frac{(1411)(79)}{(178)(1486)} =$ `r round((1411*79)/(178*1486), digits = 2)`


## _(Q2.b)_
## _(Q2.c)_
## _(Q2.d)_
## _(Q2.e)_
