---
title: "Homework 06"
author: "Spencer Pease"
date: "2/28/2020"
output: pdf_document
---

```{r setup, include=FALSE}

# Set output options
knitr::opts_chunk$set(echo = FALSE)
options(knitr.kable.NA = '-')

# Include libraries
library(dplyr)
library(readr)
library(ggplot2)
library(uwIntroStats)
library(aod)

# Helper functions
source("functions/summary_table.R")
source("functions/inference_table.R")

# Prepare data
wcgs <- read_csv("data/wcgs.csv") %>%
  transmute(
    age,
    chd = if_else(chd69 == "no", 0, 1),
    type = factor(dibpat, levels = c("A1,A2", "B3,B4"), labels = c("A", "B"))
  )

```

# _(Q1)_ Association between age and CHD events

```{r}
chd_age_lrm <- regress("odds", chd ~ age, data = wcgs)
chd_age_coef <- chd_age_lrm$fit$coefficients
```


## _(Q1.a)_

Fitted logistic regression model with _having a CHD event_ as the binary
response and _age_ and the predictor:

$$
logit(P(chd = 1 \mid age)) = \beta_0 + \beta_1 \cdot age
$$

```{r}
chd_age_inf <- chd_age_lrm %>%
  inference_table(include_intercept = TRUE) %>%
  mutate(`Pr(>|z|)` = format(`Pr(>|z|)`, digits = 3))

q1a <- list(
  chd_age_inf[1, 4:6],
  chd_age_inf[2, 4:6]
) %>% purrr::map(as.numeric) %>% purrr::map(round, digits = 2)

knitr::kable(
  chd_age_inf, booktabs = TRUE, digits = 3,
  caption = "Inferential statistics of log-odds CHD event vs age"
)

```

From the fitted model, we estimate that on average the odds of having a CHD
event within five years of the study start date for the subset of the population
zero years old ($e^{\beta_0}$) is `r q1a[[1]][1]`
(95% CI: `r q1a[[1]][2]`, `r q1a[[1]][3]`). This value is nonsensical, since it
is far outside the range of observed data and we have no reason to think the
relationship between middle-aged men and CHD events would apply to children.
Comparing two groups differing by one year in age, on average the odds ratio
$e^{\beta_1}$  of experiencing a CHD event within five years of the study start
date is `r q1a[[2]][1]` (95% CI: `r q1a[[2]][2]`, `r q1a[[2]][3]`). This is of
scientific interest for testing for an association between age and CHD events in
the population within the range of observed data.

## _(Q1.b,c)_

```{r}
wcgs_aug <- wcgs %>%
  mutate(
    log_odds_chd = chd_age_coef[1] + age * chd_age_coef[2],
    prob_chd = exp(log_odds_chd) / (1 + exp(log_odds_chd))
  )

chd_age_plot_lo <- ggplot(wcgs_aug, aes(x = age, y = log_odds_chd)) +
  geom_point() +
  labs(
    title = "Log-Odds CHD event vs Age",
    x = "Age (years)",
    y = "Log-Odds CHD event"
  ) +
  theme_bw() +
  theme(text = element_text(family = "serif"))

chd_age_plot_pr <- ggplot(wcgs_aug, aes(x = age, y = prob_chd)) +
  geom_point() +
  labs(
    title = "Probability of CHD event vs Age",
    x = "Age (years)",
    y = "Probability of CHD event"
  ) +
  theme_bw() +
  theme(text = element_text(family = "serif"))
```

```{r fig.height=4.2}
chd_age_plot_lo
chd_age_plot_pr
```

## _(Q1.d)_

```{r}
chd_age_or_rr_tbl <- tibble(
  Age = seq(40, 60, 5),
  log_odds = chd_age_coef[1] + Age * chd_age_coef[2],
  Odds = exp(log_odds),
  `Prob.` = Odds / (1 + Odds),
  OR = Odds / lag(Odds),
  RR = `Prob.` / lag(`Prob.`)
)
```

```{r}
chd_age_or_rr_tbl %>%
  select(Age, `Prob.`, Odds, RR, OR) %>%
  knitr::kable(
    booktabs = TRUE, digits = 2,
    caption = "Calculated risk ratio (RR) and odds ratio (OR)"
  )
```

From the table, we see that both the probability and odds of having a CHD event
are estimated to be higher on average for older ages. When looking at the
risk ratio (RR) and odds ratio (OR) between ages, we see that RR is lower
between older ages when the difference in age is constant, while the OR is the
same between any two ages of equal separation.

## _(Q1.e)_

```{r}
crit_val <- qnorm((1 - .95) / 2, lower.tail = FALSE)
chd_age_or5 <- (chd_age_inf[2, 2:3] * 5) %>%
  mutate(
    `95%L` = `Estimate` - crit_val * `Robust SE`,
    `95%H` = `Estimate` + crit_val * `Robust SE`
  ) %>%
  select(-`Robust SE`) %>%
  exp()
```


$$
OR_5 = \frac{Odds_{x+5}}{Odds_x}
= \frac{e^{\beta_0} e^{\beta_1 (x+5)}}{e^{\beta_0} e^{\beta_1 (x)}}
= e^{\beta_1 (x+5) - \beta_1 (x)}
= e^{\beta_1 (5)}
$$

```{r}
q1e <- list(chd_age_or5) %>%
  purrr::map(as.numeric) %>%
  purrr::map(round, digits = 2)

knitr::kable(
  chd_age_or5, booktabs = TRUE, digits = 2,
  caption = "Odds ratio comparing men 5 years apart in age"
)
```

The estimated average odds ratio between two subsets of the population five
years apart is `r q1e[[1]][1]` (95% CI: `r q1e[[1]][2]`, `r q1e[[1]][3]`). This
odds ratio matches the calculated odds ratios from (Q1.d), confirming that the
odds ratio is constant for a given interval between two groups, regardless of
the absolute value of the groups.

## _(Q1.f)_

```{r}
chd_age_inf[2, c(2:3, 7)] %>%
  knitr::kable(
    booktabs = TRUE, digits = 3,
    caption = "Wald test of age not being associated with CHD events"
  )
```

Performing a Wald test on our logistic regression model fit with robust standard
error estimates, we get a _P_-value of `r chd_age_inf[[2,7]]`.
Combined with an estimated $\beta_1 \neq 0$, we have sufficient evidence to
reject the null hypothesis that age is not associated with CHD events in the
population.

## _(Q1.g)_

```{r warning=FALSE}
lrt_test <-
  aod::anova(glm(chd~1, data = wcgs), glm(chd~age, data = wcgs), test = "LRT") %>%
  broom::tidy() %>%
  setNames(c("Resid. Df", "Resid. Dev", "df", "Deviance", "Pr(>Chi)")) %>%
  mutate(
    model = c("chd ~ 1", "chd ~ age"),
    `Pr(>Chi)` = format(`Pr(>Chi)`, digits = 3)
  ) %>%
  select(model, everything())

lrt_test[1, 6] <- NA

knitr::kable(
  lrt_test, booktabs = TRUE, digits = 1,
  caption = "likelihood ratio test of age not being associated with CHD events"
)
```

Performing a likelihood ratio test also provides sufficient evidence to reject
the null hypothesis that age is not associated with CHD events. Comparing the
_P_-value from this test to the Wald test, we see that they are similar (on the
same order of magnitude), which shows that while both tests are not equivalent,
they become more similar when working with a large number of observations.


# _(Q2)_ Association between behavior and CHD events

## _(Q2.a)_

```{r}
chd_type_tbl <- xtabs(~ chd + type, data = wcgs) %>%
  addmargins() %>%
  as_tibble() %>%
  tidyr::pivot_wider(names_from = "type", values_from = "n")

chd_type_tbl %>%
  knitr::kable(booktabs = TRUE,
    caption = "Contingency table of CHD events and behavior types"
  )
```


```{r}
chisq.test(wcgs$chd, wcgs$type) %>%
  broom::tidy() %>%
  select(`X-squared` = statistic, `P-value` = p.value, df = parameter) %>%
  mutate(`P-value` = format(`P-value`, digits = 3)) %>%
  knitr::kable(
    booktabs = TRUE, digits = 3,
    caption = "Chi-square test of CHD events and behavior types"
  )
```

With a $\chi^2$-test statistic not equal to zero and a _P_-value much less than
$0.05$, we have sufficient evidence to reject the null hypothesis that there is
no association between CHD events and behavior type in the population.

**Estimating Probability:**

 - $P(chd \mid type=A)$ : $178 \div 1589 =$ `r round(178/1589, digits = 2)`
 - $P(chd \mid type=B)$ : $79 \div 1565 =$ `r round(79/1565, digits = 2)`

**Estimating Odds:**

 - $odds(chd \mid type=A)$ : $178 \div 1411 =$ `r round(178/1411, digits = 2)`
 - $odds(chd \mid type=B)$ : $79 \div 1486 =$ `r round(79/1486, digits = 2)`

**Estimating Odds Ratio:**

 - $OR$ : $(1411 \cdot 79) \div (178 \cdot 1486) =$ `r round((1411*79)/(178*1486), digits = 2)`

## _(Q2.b)_

```{r}
chd_type_lrm <- regress("odds", chd ~ type, data = wcgs)
chd_type_coef <- chd_type_lrm$fit$coefficients

chd_type_inf <- chd_type_lrm %>%
  inference_table(include_intercept = TRUE) %>%
  mutate(`Pr(>|z|)` = format(`Pr(>|z|)`, digits = 3))

q2b <- list(
  chd_type_inf[2, 4:6] %>% as.numeric() %>% round(digits = 2),
  chd_type_inf[[2, 7]]
)

knitr::kable(
  chd_type_inf, booktabs = TRUE, digits = 3,
  caption = "Inferential statistics of LRM with response CHD and predictor Type"
)
```

From the logistic regression model fit with robust standard error estimates, we
estimate that the odds of having a CHD event given type _B_ behavior are on
average `r q2b[[1]][1]` (95% CI: `r q2b[[1]][2]`, `r q2b[[1]][3]`) times higher
than the odds of having a CHD event given type _A_ behavior in the population.
A _P_-value of `r q2b[[2]]` suggests we can reject the null hypothesis that
there is no association between CHD events and behavior type.

## _(Q2.c)_

```{r}
type_chd_lrm <- regress("odds", type ~ chd, data = wcgs)
type_chd_coef <- type_chd_lrm$fit$coefficients

type_chd_inf <- type_chd_lrm %>%
  inference_table(include_intercept = TRUE) %>%
  mutate(`Pr(>|z|)` = format(`Pr(>|z|)`, digits = 3))

knitr::kable(
  type_chd_inf, booktabs = TRUE, digits = 3,
  caption = "Inferential statistics of LRM with response Type and predictor CHD"
)
```

## _(Q2.d)_

The odds ratios ($e^{\beta_1}$) in _Q2.a_ and _Q2.b_ are the same, since they
both capture the ratio of odds of a CHD event given each behavior type. The
_P_-values are different, because each question is performing a different kind
of test of association.

## _(Q2.e)_

The odds ratios ($e^{\beta_1}$) in _Q2.b_ and _Q2.c_ are the same because odds
ratios are symmetric when swapping binary predictor and binary response (as seen
in _homework 5 Q4_).
