---
title: "Homework 02"
author: "Spencer Pease"
date: "1/22/2020"
output: pdf_document
---

```{r setup, include=FALSE}

# Set output options
knitr::opts_chunk$set(echo = FALSE)
options(knitr.kable.NA = '-')

# Include libraries
library(dplyr)
library(readr)
library(ggplot2)
library(uwIntroStats)

# Helper functions
source("functions/summary_table.R")

# Prepare data
mri <- read_table2("data/mri.txt") %>% select(age, atrophy)
fev <- read_csv("data/FEVdata.csv") %>% transmute(fev = FEV, smoke = 2 - SMOKE)

```

# _(Q1)_ MRI - Descriptive Statistics

```{r}
mri_summary <- summary_table(mri)

knitr::kable(
  mri_summary, booktabs = TRUE, digits = 3,
  caption = "Summary statistics for age and atrophy score distributions"
)

```

These summary statistics give us some basic insight into the distribution of
age and atrophy score. From the mean and median, we can tell both distributions
have a slight right skew, but are approximately centered between the first and
third quartile. Along with the IQR, the standard deviation indicates that
atrophy score has a wider distribution than age, though this does not give
insight into their relationship becuase of the incomparable units. To
understand how these two distributions relate to each other, we can also look
at a scatter plot of the data:

```{r}

mri_scatter <- ggplot(mri, aes(x = age, y = atrophy)) +
  geom_point(size = 2, alpha = .4) +
  labs(
    title = "Atrophy vs Age",
    x = "Age (years)",
    y = "Atrophy Score (unit value)"
  ) +
  theme_bw() +
  theme(text = element_text(family = "serif"))

mri_scatter
```

# _(Q2)_ MRI - Linear Regression of Mean

```{r}
mri_lm <- regress("mean", atrophy ~ age, data = mri)
```

$$ E[atrophy | age] = \hat{\beta_{0}} + \hat{\beta_{1}} \times age $$

## _(Q2.a)_
## _(Q2.b)_
## _(Q2.c)_

# _(Q3)_ MRI - Linear Regression of Geometric Mean

```{r}
mri_lm_geo <- regress("geometric mean", atrophy ~ age, data = mri)
```

## _(Q3.a)_
## _(Q3.b)_
## _(Q3.c)_

# _(Q4)_ Sensible Units

## _(Q4.a)_
## _(Q4.b)_

# _(Q5)_ Scaling Variables

# _(Q6)_ FEV - t-Test vs Regression Analysis

```{r}


fev_tt_ev <- t.test(fev ~ smoke, data = fev, var.equal = TRUE)
fev_tt_nv <- t.test(fev ~ smoke, data = fev, var.equal = FALSE)
fev_lm_ev <- lm(fev ~ smoke, data = fev)
fev_lm_nv <- regress("mean", fev ~ smoke, data = fev)

```


## _(Q6.a-f)_

# _(Q7)_ FEV - Validity of Analysis
