---
title: "MESA Analysis: C-Reactive Protein and Left Ventricular Hypertrophy"
author: "Group 3"
output: pdf_document
fontsize: 11pt
geometry: margin=.8in
indent: true
header-includes:
  - \usepackage{setspace}
  - \doublespacing
abstract: "\\singlespacing Clinical studies have shown that left ventricular mass (LVM) is a strong subclinical predictor of cardiovascular disease (CVD), which is a major cause of death globally. To characterize and investigate the relationship between inflammation and LVM, cross-sectional data from the MESA cohort was analyzed using multiple linear regression modeling. Using inflammation as a predictor variable, measured by CRP levels, and adjusting for sex and BMI, estimated mean LVM is 0.35g (95% CI: -0.90, 0.21) lower on average for each unit increase in $log_2(CRP)$. Conducting MLR adjusting for age, sex, race, BMI and systolic blood pressure (SBP), the estimated mean LVM is 2.48g (95% CI: -6.81, 1.85) lower on average for each unit increase in $log_2(CRP)$. However, both MLR models have two-sided _p_-values >0.05, which indicates insufficient evidence to reject the null hypothesis of no association between LVM and $log_2(CRP)$. The results of this study show the relationship between inflammation and LVM arises from complex mechanisms and that further research is needed to determine appropriate predictive models to aid early detection and clinical counselling for management of CVD."
---

```{r setup, include=FALSE}

# Set output options
knitr::opts_chunk$set(echo = FALSE, warning = FALSE)
options(knitr.kable.NA = '-')

# Include libraries
library(dplyr)
library(readr)
library(forcats)
library(stringr)
library(ggplot2)
library(uwIntroStats)

# Helper functions
source("functions/summary_table.R")
source("functions/inference_table.R")
source("functions/factor_inf_table.R")

# Prepare data
mesa <- read_csv("data/mesa.csv", col_types = cols(.default = col_double())) %>%
  transmute(
    `LV-mass` = lvmass,
    CRP = crp1,
    log2_CRP = log2(crp1),
    BMI = bmi1c,
    Sex = factor(gender1, levels = 0:1, labels = c("Female", "Male")),
    Race = factor(race1c, levels = 1:4, labels = c("White", "Chinese", "Black", "Hispanic")),
    Age = age,
    SBP = sbp1c
  ) %>%
  mutate_if(is.factor, fct_explicit_na)

```

## Introduction

Cardiovascular disease (CVD) is a leading cause of death, accounting for
approximately 17.9 million deaths in 2016 and 31% of global deaths (1). Clinical
research, such as the Framingham Study, have shown left ventricular hypertrophy
(LVH) is an important predictor of adverse cardiovascular events (2). In turn,
LVH is associated with multiple factors such as age, weight, race, and blood
pressure, which potentially remodel ventricular thickness and size (1-3). The
Multi-Ethnic Study of Atherosclerosis (MESA) used magnetic resonance imaging
(MRI) measurement tools to evaluate the prevalence and progression of
subclinical CVD in a cohort of 6,814 men and women between 45 and 84 years of
age across 6 sites in the US (3). Risk factors for CVD were measured, including
C-reactive protein (CRP), which is a protein used as a biomarker for
inflammation and has been associated with higher risk of cardiovascular disease
(4-5). Our first aim is to use cross-sectional data from the MESA study to
characterize the association between CRP and left ventricular mass (LVM). Our
second aim is to further investigate the complex social, demographic,
biological, and behavioral mechanisms through which inflammation contributes to
LVH, thus identifying factors and enhancing interventions for averting global
CVD deaths.


## Methods

For both aims, CRP is the predictor of interest and LVM is the response
variable. LVM was chosen as an indicator of LVH preferentially over ejection
fraction (LVEH), since LVEH is not always a good indicator of LVH, such as in
cases of dilated cardiomyopathy. Models for Aims 1 and 2 are, respectively,
shown:

$$
E(LVM \mid crp, sex, bmi) = \beta_0 + \beta_1(crp) + \beta_2(sex) + \beta_3(bmi)
$$

$$
\begin{aligned}
E(LVM \mid crp, sex, bmi, age, sbp, race) &= \beta_0 + \beta_1(crp)
+ \beta_2(sex) + \beta_3(bmi) + \beta_4(age) + \beta_5(sbp) + \beta_{6-8}(race) \\
&+ crp[\beta_9(sex) + \beta_{10}(bmi) + \beta_{11}(age) + \beta_{12}(sbp) + \beta_{13-15}(race)]
\end{aligned}
$$

The Aim 1 model is adjusted for the covariates, sex and BMI, which are already
known a priori to be associated with LVM. BMI was chosen as a predictor over
height or weight, since BMI takes into account both height and weight, and it is
already known a priori that larger people are expected to have higher LVM. The
Aim 2 model additionally includes the covariates, race, age, and systolic blood
pressure (SBP), because race, age and SBP are considered to be scientifically
plausible confounders that may distort the measure of the relationship between
CRP and LVM, as race, age, and SBP can influence both CRP and LVM individually
and are not on a causal pathway between CRP and LVH. Both linear models were fit
using robust standard error estimates, to relax the assumption of
homoscedasticity. We removed observations from our sample which included
variables that had missing values, after which our sample size remained
sufficiently large enough to assume asymptotically normal distribution of data.
The distribution of CRP is heavily right skewed, so in our models we use the
$log_2$ transformation of CRP. All statistical significance tests are performed
at the $\alpha = 0.05$ level. Our methods assumed independent data.


## Results

Baseline characteristics of $6,714$ participants are shown in Table 1. The
mean and standard deviation (SD) age was $62 \pm 10$ years and $53\%$ were female.

```{r}
stbl_cont <- mesa %>%
  mutate_if(is.factor, as.numeric) %>%
  mutate(Sex = Sex -1) %>%
  select(-Race, -Sex, everything(), Sex =, Race) %>%
  summary_table()

stbl_cont[7, 5:7] <- NA
stbl_cont[8, 4:7] <- NA

stbl_cat <- mesa %>%
  group_by(Sex, Race) %>%
  summarise(Count = n()) %>%
  ungroup() %>%
  mutate(Sex = c("Female", "", "", "", "Male", "", "", ""))

knitr::kable(list(stbl_cont, stbl_cat), booktabs = TRUE, digits = 2,
  caption = "Summary statistics (included counts of categorical variables)"
)

```

```{r}
crude_lm <- mesa %>%
  filter(!is.na(`LV-mass`)) %>%
  mutate(Sex = as.numeric(Sex) -1) %>%
  regress("mean", `LV-mass` ~ log2_CRP + Sex + BMI, data = .)

crude_inf <- crude_lm %>%
  inference_table(
    paramter_names = c("Intercept", "log2 crp", "Sex", "BMI"),
    include_intercept = TRUE
  ) %>%
  mutate(`Pr(>|t|)` = format(`Pr(>|t|)`, digits = 3))
```


In Aim 1, based on a MLR model adjusted for sex and BMI,
we found that the estimated mean LVM was lower by $0.35$g ($95\%$ CI: $-0.90$,
$0.21$) for each unit increase in $log_2(CRP)$, on average. A two-sided p-value
of $0.22$ suggests insufficient evidence to reject the null hypothesis of no
association between LVM and $log_2(CRP)$.

In exploratory analysis, a scatterplot with fitted regression lines was used to
evaluate the relationship between LVM and $log_2(CRP)$. An association between
LVM and $log_2(CRP)$ was not readily apparent among the combined sex cohort, but
when stratified by sex, a first-order trend of positive correlation between
$log_2(CRP)$ and LVM was evident among both sexes (Figure 1), suggesting that
sex is likely a confounder of the LVM-$log_2(CRP)$ relationship.

```{r}
crp_lv_sex_scatter <- ggplot(mesa, aes(x = log2_CRP, y = `LV-mass`)) +
  geom_point(aes(color = Sex), shape = 1, alpha = .3, size = .8) +
  stat_smooth(aes(color = Sex), method = "lm", se = FALSE) +
  stat_smooth(method = "lm", se = FALSE, linetype = "dashed", color = "black") +
  labs(
    title = "LV-mass vs log2 CRP",
    caption = "(Fig. 1) Colored sex−specific regression lines; dashed both−sex regression line",
    x = "log2 CRP (mg/l)",
    y = "LV-mass (g)",
    color = "Sex") +
  theme_bw() +
  theme(text = element_text(family = "serif"), legend.position = "top")

ggsave(
  "plots/crp_vs_lvmass_by_sex.png", crp_lv_sex_scatter,
  width = 7, height = 5, units = "in", dpi = 300
)
```

```{r fig.height=4.5}
crp_lv_sex_scatter
```

The Aim 2 model adjusted sex, BMI, age, race, and SBP, for confounding as well
as possible interactions with $log_2(CRP)$ for LVM. We estimated that mean LVM was
lower on average by 2.48g (95% CI: -6.81, 1.85; Table 2) for each unit increase
in $log_2(CRP)$, adjusted for sex, BMI, age, race, and SBP. A two-sided p-value of
0.26 suggests insufficient evidence to reject the null hypothesis of no
association between LVM and CRP. Among males, the association between LVM and
$log_2(CRP)$ was 1.38g (95% CI: 0.37, 2.39) higher than the female group for each 1
unit increase in $log_2(CRP)$. Furthermore, the association between LVM and
$log_2(CRP)$ was 0.13g (95% CI: -0.23, -0.02) lower for each 1 kg/m2 higher BMI.
While we found the interactions between sex, BMI and $log_2(CRP)$ to be
statistically significant (p=0.0076 and 0.015, respectively; Table 2), we failed
to find statistically significant interactions between age, race, SBP and
$log_2(CRP)$ (p=0.22, 0.23 and 0.31, respectively).

```{r}
full_lm_int <- mesa %>%
  mutate(Sex = as.numeric(Sex) -1) %>%
  filter(!is.na(`LV-mass`), !is.na(SBP)) %>%
  regress(
    "mean",
    `LV-mass` ~ log2_CRP * (Sex + BMI + Age + SBP + factor(Race)),
    data = .
  )
```

```{r}
full_lm_int_inf_tbl <- full_lm_int %>%
  factor_inf_table() %>%
  mutate(
    `Pr(>F)` = format(`Pr(>F)`, digits = 2),
    param = str_replace_all(param, c("factor\\(Race\\)" = "Race-", "-$" = "")),
    param = if_else(str_detect(param, "-"), paste0("| ", param), param)
  )

knitr::kable(full_lm_int_inf_tbl, booktabs = TRUE, digits = 2,
  caption = "Fully adjusted model"
)
```

Additionally, the LVM-$log_2(CRP)$ relationship stratified by race was examined
in a scatterplot with corresponding fitted regression lines in an exploratory
analysis. Overall, the race adjustment estimate was not statistically significant,
suggesting race does not appear to modify the LVM-$log_2(CRP)$ association.


## Discussion

While our analysis did not demonstrate statistically significant evidence for an
association between CRP and LVH, it is important to consider several important
limitations of our study. Although MESA was a prospective cohort study, we only
used cross-sectional data in our analysis, which means that causal inference
cannot be made, but instead only the strength of associations can be evaluated.
Furthermore, BMI does not distinguish between body fat vs. muscle composition,
limiting its usefulness. There is also substantial missing data among our
observations, which decreases the power of our study. For example, only 73% of
participants had an MRI of their heart, and 7% of study participants were unable
to undergo MRI due to metal within their body. This raises concern for potential
selection bias in our data, with this segment of the population that may be in
poorer health not being adequately represented.

Despite such limitations, our exploratory analyses suggest that sex is a
confounder of the association between CRP and LVH, which is biologically
plausible, given the multifactorial role of sex hormones on various organ
systems in the body, which can directly affect CRP levels and affect LVM. Our
MLR analyses suggest BMI is an effect modifier of the association between CRP
and LVH, which is also biologically plausible, given that higher BMI is often
associated with greater adiposity, a hormone-releasing tissue known to be
generally associated with higher levels of inflammation. However, it is likely
that LVH is also associated with complex mechanisms involving multiple social,
demographic, biological, and behavioral risk factors that are beyond the scope
of our data analysis, and more studies are needed to continue to advance our
understanding of the association that such variables have on LVH, which in turn
could help focus interventions to reduce CVD related morbidity and mortality
globally.


### References

 1. World Health Organization. (2013). Global action plan for the prevention and control of noncommunicable diseases 2013-2020. World Health Organization.
 2. Bluemke, D. A., et al. (2008). The relationship of left ventricular mass and geometry to incident cardiovascular events: the MESA (Multi-Ethnic Study of Atherosclerosis) study. Journal of the American College of Cardiology, 52(25), 2148-2155.
 3. Bild, D. E., et al. (2002). Multi-ethnic study of atherosclerosis: objectives and design. American journal of epidemiology, 156(9), 871-881.
 4. Lakoski, S. G., Cushman, M., Criqui, M., Rundek, T., Blumenthal, R. S., D'Agostino Jr, R. B., & Herrington, D. M. (2006). Gender and C-reactive protein: data from the Multiethnic Study of Atherosclerosis (MESA) cohort. American heart journal, 152(3), 593-598.
 5. Lakoski, S. G., Cushman, M., Palmas, W., Blumenthal, R., D’Agostino, R. B., & Herrington, D. M. (2005). The relationship between blood pressure and C-reactive protein in the Multi-Ethnic Study of Atherosclerosis (MESA). Journal of the American college of cardiology, 46(10), 1869-1874.
