---
title: "Homework 08"
author: "Spencer Pease"
date: "3/13/2020"
output: pdf_document
---

```{r setup, include=FALSE}

# Set output options
knitr::opts_chunk$set(echo = FALSE)
options(knitr.kable.NA = '-')

# Include libraries
library(dplyr)
library(readr)
library(haven)
library(ggplot2)
library(survival)
library(survminer)

# Helper functions

# Prepare data
tumor <- read_table2("data/TUMORdata.txt") %>% mutate(GX = 2 - group)
pbc <- read_dta("data/pbc.dta")

```

# _(Q1)_ Bladder Cancer Survival

```{r}
tumor_surv <- with(tumor, Surv(time, event))
tumor_fit <- survfit(tumor_surv ~ group, data = tumor)

tumor_plot <- ggsurvplot(
  tumor_fit,
  data = tumor,
  legend.labs = c("Placebo", "Thiotepa")
)

tumor_plot$plot <- tumor_plot$plot +
  theme(text = element_text(family = "serif")) +
  labs(
    title = "Kaplan-Meier Survival Curve of Tumor Relapse",
    subtitle = "Stratified by treatment group",
    x = "Time (months)",
    y = "Survival Probability"
  )

```

```{r}
tumor_plot$plot
```

```{r}
tumor_logrank <- survdiff(tumor_surv ~ group, data = tumor, rho = 0)

tumor_logrank_tbl <- tumor_logrank %>%
  broom::glance() %>%
  rename(`Chi-Sq` = statistic, `P-value` = p.value)

knitr::kable(tumor_logrank_tbl, booktabs = TRUE, digits = 3,
  caption = "Logrank test of Kaplan-Meier Curve by treatment group"
)
```

A logrank test greater than 0 shows there is some difference in the observed
vs expected values of remission by treatment group. A _p_-value over .05
suggests this difference is not statistically significant.

# _(Q2)_

## _(Q2.a)_

```{r}
tumor_cox <- coxph(tumor_surv ~ group, data = tumor)

tumor_cox_tbl <- tumor_cox %>%
  broom::tidy() %>%
  transmute(
    Predictor = term,
    coef = estimate,
    `exp(coef)` = exp(estimate),
    SE = std.error,
    `z-stat` = statistic,
    `Pr(>|z|)` = p.value,
    `exp(95%L)` = exp(conf.low),
    `exp(95%H)` = exp(conf.high)
  )

q2a <- list(
  tumor_cox_tbl[, c(3, 7, 8)] %>% as.numeric() %>% round(digits = 3),
  tumor_cox_tbl[[1, 6]] %>% round(digits = 3)
)

knitr::kable(tumor_cox_tbl, booktabs = TRUE, digits = 3,
  caption = "Summary of Cox-PH model with predictor 'group'"
)

```

The subset of the population in the thiotepa treatment group is associated with
being `r q2a[[1]][1]` (95% CI: `r q2a[[1]][2]`, `r q2a[[1]][3]`) times as
likely to come out of remission, on average, as the subset of the population in
the placebo treatment group. A _p_-value of `r q2a[[2]][1]` suggests this
difference is not statistically significant.

## _(Q2.b)_

```{r}
q2b <- list(
  tumor_logrank_tbl[["P-value"]],
  tumor_cox_tbl[["Pr(>|z|)"]]
) %>% purrr::map(~round(.x, digits = 3))

```

 - Cox-PH _p_-value: `r q2b[[1]]`
 - Kaplan-Meier Curve logrank test _p_-value: `r q2b[[2]]`

Both of these _p_-values are on the same order of magnitude and indicate that
their results are not statistically significant. The difference comes from what
the two methods are testing. The logrank test is only testing for a difference
in expected vs observed value between groups, while the Cox-PH regression tests
for an association between the response and predictor. In this case, since both
the response and predictor are binary variables, the two test are getting at a
similar relationship in different ways.

# _(Q3)_

The difference between using _group_ and _GX_ as predictors in the model, since
they both look at the treatment status of the patients, is the implicit ordering
of binary variable. _GX_ encodes “placebo” as the higher treatment group, while
_group_ encodes “thiotepa” as the higher treatment group. Since all that’s
changing is the direction, the parameter estimate $\beta_1$ will change sign
between the two groups, and the respective hazard ratios will be inverses of
each other.

```{r}
tumor_gx_cox <- coxph(tumor_surv ~ GX, data = tumor)

tumor_gx_cox_tbl <- tumor_gx_cox %>%
  broom::tidy() %>%
  transmute(
    Predictor = term,
    coef = estimate,
    `exp(coef)` = exp(estimate),
    SE = std.error,
    `z-stat` = statistic,
    `Pr(>|z|)` = p.value,
    `exp(95%L)` = exp(conf.low),
    `exp(95%H)` = exp(conf.high)
  )

q3 <- list(
  tumor_gx_cox_tbl[, c(3, 7, 8)] %>% as.numeric() %>% round(digits = 3),
  tumor_gx_cox_tbl[[1, 6]] %>% round(digits = 3)
)

knitr::kable(tumor_gx_cox_tbl, booktabs = TRUE, digits = 3,
  caption = "Summary of Cox-PH model with predictor 'GX'"
)

```

# _(Q4)_ Primary Biliary Cirrhosis Survival

```{r}
pbc_surv <- with(pbc, Surv(years, status))
pbc_cox <- coxph(pbc_surv ~ albumin, data = pbc)

pbc_cox_tbl <- pbc_cox %>%
  broom::tidy() %>%
  transmute(
    Predictor = term,
    coef = estimate,
    `exp(coef)` = exp(estimate),
    SE = std.error,
    `z-stat` = statistic,
    `Pr(>|z|)` = p.value,
    `exp(95%L)` = exp(conf.low),
    `exp(95%H)` = exp(conf.high)
  ) %>%
  mutate(`Pr(>|z|)` = format(`Pr(>|z|)`, digits = 3))

q4 <- list(
  pbc_cox_tbl[, c(3, 7, 8)] %>% as.numeric() %>% round(digits = 3),
  pbc_cox_tbl[[1, 6]]
)

knitr::kable(pbc_cox_tbl, booktabs = TRUE, digits = 3,
  caption = "Summary of Cox-PH model with predictor 'albumin'"
)
```

Two subsets of the population differing by $1 \frac{g}{dl}$ albumin are
associated with hazards of death `r q4[[1]][1]` (95% CI: `r q4[[1]][2]`,
`r q4[[1]][3]`) times as high, on average, in the higher group. A _p_-value of
`r q4[[2]][1]` suggests this difference is statistically significant.

# _(Q5)_

## _(Q5.a)_

```{r}
pbc_beta <- pbc_cox$coefficients

pbc_diff_tbl <-
  tibble(
    reference = 3,
    value = c(2.5, 3.5, 4)
  ) %>%
  mutate(
    `log(hazard) diff` = pbc_beta * (value - reference),
    `hazard ratio` = exp(`log(hazard) diff`)
  )

knitr::kable(pbc_diff_tbl, booktabs = TRUE, digits = 3,
  caption = "Estimated log(hazard) difference and HR for differences in 'albumin'"
)

```

## _(Q5.b,c)_

The hazard ratio for $2.5 \frac{g}{dl}$ to $3 \frac{g}{dl}$ is the inverse of
the hazard ratio for $3.5 \frac{g}{dl}$ to $3 \frac{g}{dl}$.

Since the difference between $3.5$ and $2.5$ is $1$, the estimated hazard ratio
would be the same as the exponentiated parameter estimate ($e^{\beta_1}$).

# _(Q6)_ CVD Smoking and Death

## _(Q6.a)_

We should adjust for age in this analysis, since we are interested in the
association between smoking and death. It’s a given that older people are more
likely to die for many reasons, but we are interested in the direct effect
smoking has on mortality. Adjusting for age allows us to get at this
relationship, controlling for the additional mortality associated with age.

## _(Q6.b)_

Since there are significantly more people in the sample who don’t smoke, and
they are on average older than those who do smoke, we would expect the age
adjusted hazard ratio for smoking to be higher than the unadjusted hazard ratio.
Older people are generally more likely to die, so the age-related deaths in the
higher ages made the smoking-related deaths in the younger ages more similar,
hiding the real hazard of smoking. Adjusting for age shows the direct hazard of
smoking.

## _(Q6.c)_

The 95% confidence interval is calculated from the estimated coefficient, then
exponentiated, and not from the exponentiated coefficient directly. This means
we can’t expect the confidence interval to be symmetric around the estimated
hazard ratio, since they are not calculated from each other.
