---
title: "Homework 05"
author: "Spencer Pease"
date: "2/14/2020"
output: pdf_document
---

```{r setup, include=FALSE}

# Set output options
knitr::opts_chunk$set(echo = FALSE)
options(knitr.kable.NA = '-')

# Include libraries
library(dplyr)
library(readr)
library(ggplot2)
library(uwIntroStats)

# Helper functions
source("functions/summary_table.R")
source("functions/inference_table.R")

# Prepare data
mri <- read_table2("data/mri.txt") %>%
  select(age, sbp, male, race) %>%
  mutate(
    sex = factor(male, levels = c(0, 1), labels = c("Female", "Male")),
    race = factor(race, levels = c(4, 1:3), labels = c("Other", "White", "Black", "Asian")),
    race_dummy = factor(race, labels = c("Rw", "Rb", "Ra", "Ro"))
  ) %>%
  tidyr::pivot_wider(
    names_from = "race_dummy",
    values_from = "race_dummy",
    values_fn = list(race_dummy = length),
    values_fill = list(race_dummy = 0)
  ) %>%
  select(-Ro)

```

# _(Q1)_  Systolic blood pressure by age and sex

## _(Q1.a)_

```{r}
sbp_age_scatter <- ggplot(mri, aes(x = age, y = sbp, color = sex)) +
  geom_point(shape = 1, alpha = .5) +
  stat_smooth(method = "loess", se = FALSE) +
  labs(title = "SBP vs Age", x = "Age (years)", y = "SBP (mmHg)") +
  theme_bw() +
  theme(text = element_text(family = "serif"), legend.position = "top")
```

```{r}
sbp_age_scatter
```

## _(Q1.b)_

## _(Q1.c)_

```{r}
sbp_age_sex_lm <- regress("mean", sbp ~ age*male, data = mri)
sbp_age_sex_lm_inf <-
  inference_table(sbp_age_sex_lm, include_intercept = TRUE) %>%
  mutate(`Pr(>|t|)` = format(`Pr(>|t|)`, digits = 3))

knitr::kable(sbp_age_sex_lm_inf, booktabs = TRUE, digits = 3,
  caption = "Inferential statistics of SBP vs age, adjusted for sex"
)

```


## _(Q1.d)_

# _(Q2)_ Systolic blood pressure by age and race

## _(Q2.a)_

Model of race-adjusted association between mean systolic blood pressure and
age:

$$
E(sbp \mid age,race) = \beta_{0} + \beta_{1} \cdot age
+ \beta_{2-4} \cdot race
+ \beta_{5-7} \cdot age \cdot race
$$

where _race_ is treated as a categorical variable, with possible values _White_,
_Black_, _Asian_, and _Other_ (treated as the reference category).

<!--
In this model, race is encoded as:

 * $R_{w}=1$, $R_{b}=R_{a}=0$ : White
 * $R_{b}=1$, $R_{w}=R_{a}=0$ : Black
 * $R_{a}=1$, $R_{w}=R_{b}=0$ : Asian
 * $R_{w}=R_{b}=R_{a}=0$ : Other
-->

## _(Q2.b)_

```{r}
old_sbp_age_race_lm <- regress("mean", sbp ~ age + age*Rw + age*Rb + age*Ra, data = mri)

old_sbp_age_race_lm_inf <-
  inference_table(old_sbp_age_race_lm, include_intercept = TRUE) %>%
  mutate(`Pr(>|t|)` = format(`Pr(>|t|)`, digits = 3))

sbp_age_race_lm <- regress("mean", sbp ~ age * as.factor(race), data = mri)

sbp_age_race_lm_inf <- sbp_age_race_lm$model %>%
  as.table() %>%
  as_tibble(.name_repair = "minimal") %>%
  setNames(c("param", "var", "val")) %>%
  tidyr::pivot_wider(names_from = "var", values_from = "val") %>%
  mutate(Parameter = c(
    "Intercept", "age",
    "race", "| race-White", "| race-Black", "| race-Asian",
    "age:race", "| age:race-White", "| age:race-Black", "| age:race-Asian"
  )) %>%
  select(Parameter, df, everything(), -`F stat`, -param, -`Naive SE`)

knitr::kable(sbp_age_race_lm_inf, booktabs = TRUE, digits = 3,
  caption = "Inferential statistics of SBP vs age, adjusted for race"
)

```


# _(Q3)_ Systolic blood pressure by age, sex, and race

## _(Q3.a)_

Model of the association between SBP and age, adjusted for race and sex:

$$
E(sbp \mid age,male,race) = \beta_{0} + \beta_{1} \cdot age
+ \beta_{2} \cdot male
+ \beta_{3-5} \cdot race
+ (\beta_{6} \cdot male + \beta_{7-9} \cdot race) \cdot age
$$

<!--
$$
\begin{aligned}
E(sbp \mid age, race, sex) &= \beta_{0} \\
&+ \beta_{1} \cdot age \\
&+ \beta_{2} \cdot male \\
&+ \beta_{3} \cdot male \cdot age \\
&+ \beta_{4} \cdot R_{w} + \beta_{5} \cdot R_{b} + \beta_{6} \cdot R_{a} \\
&+ (\beta_{7} \cdot R_{w} + \beta_{8} \cdot R_{b} + \beta_{9} \cdot R_{a}) \cdot age
\end{aligned}
$$
-->

```{r}
old_sbp_all_lm <- regress("mean", sbp ~ age*male + age*Rw + age*Rb + age*Ra, data = mri)
old_sbp_all_lm_inf <- inference_table(old_sbp_all_lm, include_intercept = TRUE)

sbp_all_lm <- regress("mean", sbp ~ age*male + age*as.factor(race), data = mri)

sbp_all_lm_inf <- sbp_all_lm$model %>%
  as.table() %>%
  as_tibble(.name_repair = "minimal") %>%
  setNames(c("param", "var", "val")) %>%
  tidyr::pivot_wider(names_from = "var", values_from = "val") %>%
  mutate(Parameter = c(
    "Intercept", "age", "male",
    "race", "| race-White", "| race-Black", "| race-Asian",
    "age:male",
    "age:race", "| age:race-White", "| age:race-Black", "| age:race-Asian"
  )) %>%
  select(Parameter, df, everything(), -`F stat`, -param, -`Naive SE`)

knitr::kable(sbp_all_lm_inf, booktabs = TRUE, digits = 3,
  caption = "Inferential statistics of SBP vs age, adjusted for race and sex"
)
```

## _(Q3.b)_

```{r}
sbp_all_pred <- predict(
  sbp_all_lm, interval = "prediction",
  newdata = data.frame(age = 70, male = 0, race = "Black")
)

# data.frame(age = 70, male = 0, Rw = 0, Rb = 1, Ra = 0)

colnames(sbp_all_pred) <- c("Prediction", "95%L", "95%H")

knitr::kable(sbp_all_pred, booktabs = TRUE, digits = 3,
  caption = "95% PI for SBP among 70-year-old black women"
)

```

## _(Q3.c)_

```{r}
res_v_fit_data <- tibble(
  fit = sbp_all_lm$linearPredictor,
  res = sbp_all_lm$residuals
)

res_v_fit_plot <- ggplot(data = res_v_fit_data, aes(x = fit, y = res)) +
  geom_point(shape = 1, alpha = .5) +
  labs(title = "Residuals vs Fitted Values", x = "Fitted values", y = "Residuals") +
  theme_bw() +
  theme(text = element_text(family = "serif"), legend.position = "top")
```

```{r}
res_v_fit_plot
```



# _(Q4)_ Odds ratios

## _(Q4.a-c)_

Odds ratio for disease, given exposure:

$$
OR_{(D|E=1)}
= \frac{Odds_{D|E=1}}{Odds_{D|E=0}}
= \frac{P_{(D|E=1)} \div (1-P_{(D|E=1)})}{P_{(D|E=0)} \div (1-P_{(D|E=0)})}
= \frac{d/c}{b/a}
$$
Odds ratio for exposure, given disease:

$$
OR_{(E|D=1)}
= \frac{Odds_{E|D=1}}{Odds_{E|D=0}}
= \frac{P_{(E|D=1)} \div (1-P_{(E|D=1)})}{P_{(E|D=0)} \div (1-P_{(E|D=0)})}
= \frac{d/b}{c/a}
$$

Comparing these two odds ratios, we see they are identical:

$$
OR_{(D|E=1)} = \frac{d/c}{b/a} = \frac{ad}{bc} = \frac{d/b}{c/a} = OR_{(E|D=1)}
$$
